---
title: "Seroprevalence and test characteristic meta-analysis"
author: "Bob Carpenter and Andrew Gelman"
date: "May 2020"
output:
  tufte::tufte_html:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
options(digits = 3)
options("width" = 180)

library(ggplot2)

library(gridExtra)

library(knitr)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
knitr::opts_chunk$set(comment = "")

library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores(logical = FALSE))

library(tufte)

ggtheme_tufte <- function() {
  theme(plot.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       size = 0.5,
                       linetype = "solid"),
        plot.margin=unit(c(1, 1, 0.5, 0.5), "lines"),
        panel.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       size = 0.5,
                       linetype = "solid"),
        panel.grid.major = element_line(colour = "white", size = 1, linetype="dashed"),
          # blank(),
        panel.grid.minor = element_blank(),
        legend.box.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       linetype = "solid"),
        axis.ticks = element_blank(),
        axis.text = element_text(family = "Palatino", size = 16),
        axis.title.x = element_text(family = "Palatino", size = 20,
                                    margin = margin(t = 15, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(family = "Palatino", size = 18,
                                    margin = margin(t = 0, r = 15, b = 0, l = 0)),
        strip.background = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid"),
        strip.text = element_text(family = "Palatino", size = 16),
        legend.text = element_text(family = "Palatino", size = 16),
        legend.title = element_text(family = "Palatino", size = 16,
                                    margin = margin(b = 5)),
        legend.background = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid"),
        legend.key = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid")
  )
}

printf <- function(msg, ...) cat(sprintf(msg, ...))


print_file <- function(file) {
  cat(paste(readLines(file), "\n", sep=""), sep="")
}
```


# Introduction

## Sensitivity, specificity, and prevalence

Let the random variable

* $Z \in \{ 0, 1 \}$ represent the disease status of an individual,

where $Z = 1$ indicates the individual is positive for the disease and
$Z = 0$ indicates they are negative.

The *prevalence* of a disease (in a given [sub]population) is the
probability of an individual being positive,
$$
\textrm{prevalence}
= \textrm{Pr}[Z = 1].
$$

Let the random variable

* $Y \in \{ 0, 1 \}$ represent the result of a diagnostic test,

where $Y = 1$ indicates a positive test result and
$Y = 0$ a negative result.

The *sensitivity* of a test is its accuracy for positive individuals,
$$
\textrm{sensitivity}
= \textrm{Pr}[Y = 1 \mid Z = 1].
$$

The *specificity* of a test is its accuracy for negative individuals,
$$
\textrm{specificity}
= \textrm{Pr}[Y = 0 \mid Z = 0].
$$


# Data

The data (Bendavid et al. 2020) consists of 17 seroprevalence
studies

* $N^{\textrm{sens}} = 3$ sensitivty tests of subjects known to be positive,
* $N^{\textrm{spec}} = 13$ specificity tests of subjects known to be
negative, and
* $N^{\textrm{unk}} = 1$ test of subjects of unknown status.

### Sensitivity tests ($Z = 1$)

```{r, echo = FALSE}
sens_tests <- 3
sens_vals <-
c(78, 85,
  27, 37,
  25, 35)
sens_mat <- matrix(sens_vals, nrow = sens_tests, ncol = 2, byrow = TRUE)
sens_df <- data.frame(pos_tests = sens_mat[ , 1],
                      tests = sens_mat[ , 2],
		      sample_sens = sens_mat[ , 1] / sens_mat[ , 2])
knitr::kable(sens_df)
```

### Specificity tests ($Z = 0$)

```{r, echo = FALSE}
spec_tests <- 13
spec_vals <-
c(368, 371,
   30,    30,
   70,    70,
  1102, 1102,
   300,  300,
   311,  311,
   500,  500,
   198,  200,
    99,   99,
    29,   31,
   146,  150,
   105,  108,
    50,   52)
spec_mat <- matrix(spec_vals, nrow = spec_tests, ncol = 2, byrow = TRUE)
spec_df <- data.frame(neg_tests = spec_mat[ , 1],
                      tests = spec_mat[ , 2],
		      sample_spec = spec_mat[ , 1] / spec_mat[ , 2])
knitr::kable(spec_df)
```

### Prevalence tests ($Z$ unknown)

```{r, echo = FALSE}
unk_tests <- 1
unk_vals <-
c(50, 3330)
unk_mat <- matrix(unk_vals, nrow = unk_tests, ncol = 2, byrow = TRUE)
unk_df <- data.frame(pos_tests = unk_mat[ , 1],
                     tests = unk_mat[ , 2],
		     sample_prev = unk_mat[ , 1] / unk_mat[ , 2])
knitr::kable(unk_df)
```

## Pooled estimates

The pooled sample sensitivity is
`r sum(sens_mat[ , 1]) / sum(sens_mat[ , 2])`,
pooled sample specificity is
`r sum(spec_mat[ , 1]) / sum(spec_mat[ , 2])`,
and pooled sample prevalence is
`r sum(unk_mat[ , 1]) / sum(unk_mat[ , 2])`.


# Sensitivity and specificity meta-analysis

Using sensitivity as an example (specificity is identical other than
some signs), the meta-analysis assumes a test's sensitivity and
specificity vary by site with normal noise on the log odds scale.  We
will parameterize sensitivity and specificity using log odds, where
the *log odds function* $\textrm{logit}:(0, 1) \rightarrow \mathbb{R}$
is
$$
\textrm{logit}(u) = \log \frac{u}{1 - u}.
$$
Its inverse is the *logistic function*,
$\textrm{logit}^{-1}:\mathbb{R} \rightarrow (0, 1),$
$$
\textrm{logit}^{-1}(v) = \frac{1}{1 + \exp(-v)}.
$$

## Varying sensitivity and specificity

The sensitivity of site $k$ will be represented by the variable
$\theta_{1, k}$ (the $1$ is for sensitivity as opposed to
specificity).

## Sampling distribution (likelihood)

The tests are assumed to be independent so that the number of positive
tests at a site may be modeled as binomial,
$$
y_k \sim \textrm{binomial_logit}(N_k, \theta_{1, k}).
$$
We've used a parameterization of the binomial's chance of success
parameter on the log odds scale, where
$$
\textrm{binomial_logit}(N, \theta)
= \textrm{binomial}(N, \textrm{logit}^{-1}(\theta)).
$$

## Hierarchical prior

The model will assume a test sensitivity log odds $\mu_1$ and scale of
variation $\sigma_1$ such that the sensitivity log odds $\theta_{1, k}$ at site
$k$ is distributed as
$$
\theta_{1, k} \sim \textrm{normal}(\mu_1, \sigma_1)
$$
This will pool (or regularize) the estimates $\theta_{1, k}$ toward
$\mu_1$, with $\sigma_1$ controlling the scale of inter-site variation
in sensitivity.

As $\sigma_1 \rightarrow 0$, the model approaches a *full pooling
model* where we assume there is no variation in sensitivity among the
sites.  As $\sigma_1 \rightarrow \infty$, the model approaches a
*no-pooling model* where each site's sensitivity is estimated
independently. The hierarchical model is a *partial pooling model*,
which determines the amount of pooling to apply from the data by
controlling the estimate of $\sigma_1$ from the data.

## Hyperpriors

The hyperparameters $\mu_1, \sigma_1$ should get hyperpriors based
on what we know about this disease and serum testing variation among
sites.  Assuming we know nothing, we could assume that the
probability of a correct test, $\textrm{logit}^{-1}(\mu_1)$ has a
uniform distribution, which has an implied standard logistic
distribution on the log odds scale,
$$
\mu_1 \sim \textrm{logistic}(0, 1).
$$
The scale parameter should be consistent with zero, so we do not want
to give it a lognormal or inverse-gamma family prior.  We do not
expect a huge variation on the log odds scale, so we can use a simple
standard half-normal hyperprior on $\sigma_1$,
$$
\sigma_1 \sim \textrm{normal}_+(0, 1)
$$

If more is known about the tests, these hyperpriors should be
tighter.  Without a lot of data (both per site and number of sites),
the posteriors will be sensitive to these hyperpriors.

## Predictive inference for new data at existing sites

To predict future test results at site $k$, the probability of getting
$n$ positive tests out of the next $N$ tests where the subject is
positive, given observed (training) data $y$, is given by the *posterior
predictive distribution*,
\begin{eqnarray*}
p(n \mid N, k)
& = &
\int
\underbrace{\textrm{binomial_logit}(n \mid N, \theta_{1, k})}
_{\textrm{sampling uncertainty}}
\cdot
\underbrace{p(\theta_{1, k} \mid y)}
_{\textrm{estimation uncertainty}}
     \, \textrm{d}{\theta_{1, k}}
\\[6pt]
& \approx &
\frac{1}{M} \sum_{m = 1}^M
\textrm{binomial_logit}(n \mid N, \theta_{1, k}^{(m)}),
\end{eqnarray*}
where
$$
\theta_{1, k}^{(m)} \sim p(\theta_{1, k} \mid y)
$$
is a draw from the posterior (binomial-logit determines sampling
uncertainty and drawing from the posterior the estimation
uncertainty).


## Predictive inference for new data at new sites

Suppose we gather data from a new testing site $k'$ not in the
original data.  The posterior predictive distribution for new data
is identical in form,
\begin{eqnarray*}
p(n \mid N, k')
& = &
\int
\underbrace{\textrm{binomial_logit}(n \mid N, \theta_{1, k'})}
_{\begin{array}{l}
  \textrm{sampling} \\
  \textrm{uncertainty}
  \end{array}}
\cdot
\underbrace{p(\theta_{1, k'} \mid y)}
_{\begin{array}{l}
  \textrm{estimation} \\
  \textrm{uncertainty}
  \end{array}}

\, \textrm{d}{\theta_{1, k'}}
\\[6pt]
& \approx &
\frac{1}{M} \sum_{m = 1}^M
\textrm{binomial_logit}(n \mid N, \theta_{1, k'}^{(m)}),
\end{eqnarray*}
where $\theta_{1, k'}^{(m)} \sim p(\theta_{1, k'} \mid y)$ is a draw
from the posterior.

The only difference is that now there is no data informing $\theta_{1,
k'}$, only the hyperprior parameters $\mu_1, \sigma_1$ estimated
through the meta-analysis, that is, as $\theta_{1, k'} \sim
\textrm{normal}(\mu_1, \sigma_1)$.  This allows the sampling for MCMC
to be refactored as
$$
\theta_{1, k'}^{(m)} \sim \textrm{normal}(\mu_1^{(m)},
\sigma_1^{(m)}),
$$
where $\mu_1^{(m)}, \sigma_1^{(m)} \sim p(\mu_1, \sigma_1 \mid y).$


# Putting it all together to estimate diseases prevalence

Specificity meta-analysis is identical to sensitivity meta-analysis
(with some signs reversed).  We can now use the predictive inference
for sensitivity and specificity in our new site along with the
observed number of positive tests at the new site to estimate
prevalence of the disease (in the sampled (sub)population). Suppose we
have a site with sensitivity $\theta_{1, k}$ and a specificity of
$\theta_{0, k}$.  If the disease prevalence is $\pi$, the probability
of getting a positive test result is
$$
\mbox{Pr}[y = 1]
= \pi \cdot \textrm{logit}^{-1}(\theta_{1, k})
                            + (1 - \pi) \cdot
			    (1 - \textrm{logit}^{-1}(\theta_{0, k}))
$$
For a total of $N$ independent tests, the number of positive tests
will thus be distributed binomially with the preceding probability of
success,
$$
n \sim \textrm{binomial}\left(N,
\pi \cdot \textrm{logit}^{-1}(\theta_{1, k})
                            + (1 - \pi) \cdot
			    (1 - \textrm{logit}^{-1}(\theta_{0, k}))\right).
$$

If we assume a simple uniform prior on prevalence $\pi \in (0, 1)$,
the entire model is
\begin{eqnarray*}
\pi & \sim & \textrm{uniform}(0, 1)
\\
\mu_0, \mu_1 & \sim & \textrm{logistic}(0, 1)
\\
\sigma_0, \sigma_1 & \sim & \textrm{normal}_+(0, 1)
\\
\theta_{0, k} & \sim & \textrm{normal}(\mu_0, \sigma_0)
\\
\theta_{1, k} & \sim & \textrm{normal}(\mu_1, \sigma_1)
\\
n^{\textrm{pos}}_k & \sim & \textrm{binomial}(N^{\textrm{pos}}_k,
\theta_{1, k})
\\
n^{\textrm{neg}}_k & \sim & \textrm{binomial}(N^{\textrm{neg}}_k, \
\theta_{0, k})
\\
n^{\textrm{unk}}_k & \sim &
\textrm{binomial}\left(N^{\textrm{unk}}_k,
  \pi \cdot \textrm{logit}^{-1}(\theta_{1, k})
  + (1 - \pi) \cdot (1 - \textrm{logit}^{-1}(\theta_{0, k}))
\right)
\end{eqnarray*}
Some care is required to avoid clasing $k$ subscripts in an implementation.

The posterior $p(\pi \mid n^{\textrm{pos}}, N^{\textrm{pos}},
n^{\textrm{neg}}, N^{\textrm{neg}}, n^{\textrm{unk}},
N^{\textrm{unk}})$ characterizes what we know about the prevalence of
the disease given the observed sensitivity data, specificity data, and
unkown status data.

# Stan Model

Here is the full Stan model implementing this meta-analysis.

```{r, echo = FALSE}
print_file("meta.stan")
```

Now we compile it.
```{r}
model <- stan_model("meta.stan")
```
and then fit
```{r}
data <-
list(sigma_sigma_logit_sens = 0.01,
     sigma_sigma_logit_spec = 0.01,
     K_pos = sens_tests,
     N_pos = sens_df$tests,
     n_pos = sens_df$pos_tests,
     K_neg = spec_tests,
     N_neg = spec_df$tests,
     n_neg = spec_df$neg_tests,
     K_unk = unk_tests,
     N_unk = array(unk_df$tests, dim = c(length(unk_df$tests))),
     n_unk = array(unk_df$pos_tests, dim = c(length(unk_df$tests))))

fit <- sampling(model, data = data,
                iter = 4000,
		control = list(adapt_delta = 0.99),
		open_progress = FALSE, refresh = 0)

print(fit, probs = c(0.025, 0.5, 0.975), digits = 3, pars = c("pi"))
```

# References

* Gelman, Andrew.  1 May 2020.  [Simple Bayesian analysis inference of
  coronavirus infection rate from the Stanford study in Santa Clara
  county](https://statmodeling.stat.columbia.edu/2020/05/01/simple-bayesian-analysis-inference-of-coronavirus-infection-rate-from-the-stanford-study-in-santa-clara-county/).
  *Statistical Modeling, Causal Inference, and Social Sciences* blog.

* Eran Bendavid, Bianca Mulaney, Neeraj Sood, Soleil Shah, Emilia Ling,
Rebecca Bromley-Dulfano, Cara Lai, Zoe Weissberg, Rodrigo
Saavedra-Walker, Jim Tedrow, Dona Tversky, Andrew Bogan, Thomas
Kupiec, Daniel Eichner, Ribhav Gupta, John P.A. Ioannidis, Jay
Bhattacharya. 27 April 2020.  [COVID-19 Antibody Seroprevalence in
Santa Clara County, California, version 2](https://www.medrxiv.org/content/10.1101/2020.04.14.20062463v2.full.pdf)
